
# Cancer type Prediction from CNA Enhanced + Other Categorical Predictors

```{r}

suppressPackageStartupMessages({
library(tidyr)
library(dplyr)
library(pROC)
library(randomForest)
library(ggplot2)
library(MASS)
library(glmnet)
library(ranger)
library(doParallel)
library(foreach)
library(xgboost)
library(dbarts)
})


```

```{r}

#Profile_description: Putative copy-number from GISTIC 2.0. Values: -2 = homozygous deletion; -1 = hemizygous deletion; 0 = neutral / no change; 1 = gain; 2 = high level amplification.
#Profile_name: Putative copy-number alterations from GISTIC
data_cna = read.delim("/Users/ali/Downloads/msk_chord_2024/data_cna.txt", header = T)
data_cna = data.frame(t(data_cna))
colnames(data_cna) = data_cna[1,]
data_cna = data_cna[-1,]
row.names(data_cna) = gsub("\\.","-",row.names(data_cna))
data_cna[] <- lapply(data_cna, as.integer)
data_cna$Sample_ID = row.names(data_cna)

#-------------------------------------------------------------------------------

#Clinical data directly from cBioPortal

clinical_data_cbio = read.delim("/Users/ali/Downloads/msk_chord_2024_clinical_data-2.tsv")

```

### Custom functions

```{r}

set.seed(1010)

#Removes columns that have NAs in them - machine learning models cannot handle NAs; difficult to impute values when placeholders such as "0" might be used to singify a disease state. Imputation is difficult without making big assumptions, especially since this is beca use some test may not have been performed for certain patients. Therefore, I removed those columns entirely. 
na_cleanup_columns = function(dataframe)
{
  return(dataframe[,colSums(is.na(dataframe)) == 0])
}

#-------------------------------------------------------------------------------

#Creates training rows
train_rows = function(x)
{
  sample(1:nrow(x),nrow(x)*(2/3), replace = F)
}

#-------------------------------------------------------------------------------

#Custom function that uses glmnet to compute ROC and AUC using parallel computing for logistic regression with varying degrees of L1 and L2 penalties, essentially switching between Ridge, Lasso, and Elastic-net. 

elastic_net_logistic = function(training_x, training_y, a, folds, testing_x, testing_y, parallel)

{

if (parallel == TRUE)
{
  # Set up parallel backend
  cores = parallel::detectCores()
  cl = makeCluster(cores*(1/2))
  registerDoParallel(cl)
  
  cv_lasso_logistic = cv.glmnet(as.matrix(training_x), training_y, family = "binomial", alpha = a, nfolds = folds, parallel = TRUE)
  
  # Shut down parallel backend
  stopCluster(cl)
  registerDoSEQ()
}

else
{cv_lasso_logistic = cv.glmnet(as.matrix(training_x), training_y, family = "binomial", alpha = a, nfolds = folds)}

testing_x_regularized_logisitic = as.matrix(testing_x)
lasso_probs = predict(cv_lasso_logistic, 
                      newx = testing_x_regularized_logisitic,
                      type = "response",
                      s=cv_lasso_logistic$lambda.1se)

coefficients_glmnet = as.data.frame(as.matrix(coef(cv_lasso_logistic, s = cv_lasso_logistic$lambda.1se)))
coefficients_glmnet$Genes = row.names(coefficients_glmnet)
genes_selected = coefficients_glmnet[coefficients_glmnet$s1 != 0,][-1,] #removing the intercept

# Calculate AUC
roc_lasso = pROC::roc(testing_y, lasso_probs)

# Plot ROC Curve
roc_data = data.frame(TPR = roc_lasso$sensitivities, 
                      FPR = 1 - roc_lasso$specificities)

return(list(ROC_Data = roc_data, AUC = roc_lasso$auc, Genes = genes_selected))

}

#-------------------------------------------------------------------------------

LDA_function = function(training_data, testing_x, testing_y)

{
  
lda_model = lda(Cancer_Type ~ ., data = training_data)

# Make predictions
lda_predictions = predict(lda_model, newdata = testing_x)

# Predicted probabilities for ROC curve
lda_probs = lda_predictions$posterior[, 2]  # Probabilities for class "1"

# Calculate AUC
lda_roc = pROC::roc(testing_y, lda_probs)

# Plot ROC curve
lda_roc_data = data.frame(TPR = lda_roc$sensitivities, 
                          FPR = 1 - lda_roc$specificities)

return(list(ROC_Data = lda_roc_data, AUC = lda_roc$auc))

}

#-------------------------------------------------------------------------------

logistic_regression = function(training_data, testing_x, testing_y)
  
{

logistic_model = glm(Cancer_Type ~ ., 
                     data = training_data, 
                     family = "binomial")

predictions = predict(logistic_model, 
                      newdata = testing_x,
                      type = "response")

roc_curve = pROC::roc(testing_y, predictions)

# Extract ROC curve data
roc_data = data.frame(TPR = roc_curve$sensitivities,  # True positive rate
                      FPR = 1 - roc_curve$specificities)  # False positive rate

return(list(ROC_Data = roc_data, AUC = roc_curve$auc))

}

#-------------------------------------------------------------------------------

random_forest_ranger = function(training_data, training_y, testing_x, testing_y, num_trees, mtry = NA, cores = NA)

{
  
mtry_value = mtry

if (is.na(mtry))
{
mtry_value = floor(sqrt(ncol(training_data)))  
}

if (is.na(cores))
{
cores = 1
}

rf_model = ranger(dependent.variable.name = training_y,
                  data = training_data,
                  mtry = mtry_value,
                  num.trees = num_trees,
                  importance = "impurity",
                  probability = TRUE,
                  num.threads = cores)  # Number of cores to use

rf_probs = predict(rf_model, data = testing_x)$predictions[, 2]

rf_roc = pROC::roc(testing_y, rf_probs)

rf_roc_data = data.frame(TPR = rf_roc$sensitivities, 
                         FPR = 1 - rf_roc$specificities)

variable_importance = as.data.frame(rf_model$variable.importance)
variable_importance$Variable = rownames(variable_importance)
colnames(variable_importance) = c("Importance", "Variable")

variable_importance = variable_importance[order(variable_importance$Importance, decreasing = TRUE), ]

return(list(ROC_Data = rf_roc_data, AUC = rf_roc$auc, VarImportance = variable_importance))

}

#-------------------------------------------------------------------------------

xgboost_parallel = function(training_x, training_y, testing_x, testing_y, learning_rate = NA, cores = NA, n_tree = NA, tree_depth = NA)

{
  
if(is.na(learning_rate))
{
  learning_rate = 0.3
}

if(is.na(cores))
{
  cores = 1
}

if(is.na(n_tree))
{
  n_tree = 500
}
  
if(is.na(tree_depth))
{
  tree_depth = 6
}

# Separate features and target
X_train = as.matrix(training_x)
y_train = as.numeric(training_y) - 1  # Ensure labels are 0-based

X_test = as.matrix(testing_x)
y_test = as.numeric(testing_y) - 1  # Ensure labels are 0-based

#y_test = as.numeric(test_data_xgb$Cancer.Type) - 1  # Ensure labels are 0-based


# Define hyperparameters
xgb_params = list(objective = "binary:logistic",  # Binary classification
                  eval_metric = "auc",
                  max_depth = tree_depth,        # Tree depth
                  eta = learning_rate)           # Learning rate

# Train the XGBoost model
xgb_train = xgb.DMatrix(data = X_train, label = y_train)

xgb_model = xgb.train(params = xgb_params,
                      data = xgb_train,
                      nrounds = n_tree,
                      nthread = cores,
                      verbose = 0)

# Make predictions
xgb_test = xgb.DMatrix(data = X_test)
xgb_probs = predict(xgb_model, xgb_test)

# Calculate ROC and plot
xgb_roc = pROC::roc(y_test, xgb_probs)

xgb_roc_data = data.frame(TPR = xgb_roc$sensitivities,
                          FPR = 1 - xgb_roc$specificities)

# Variable importance
xgb_importance = xgb.importance(model = xgb_model, feature_names = colnames(X_train)) %>%
  arrange(desc(Gain))

return(list(ROC_Data = xgb_roc_data, AUC = xgb_roc$auc, VarImportance = xgb_importance))

}

#-------------------------------------------------------------------------------

factor_2_integer = function(x)
{
  (as.integer(as.factor(x)))
}

```

**Training & testing split**

```{r}

set.seed(1010)

clinical_cna_data = na_cleanup_columns(inner_join(clinical_data_cbio, data_cna, by = c("Sample.ID" = "Sample_ID")))

dim(clinical_cna_data)

#Use this code to add any other features to the dataset
clinical_cna_data_for_model = na.omit(inner_join(clinical_cna_data, 
                                                         clinical_data_cbio[,c("Sample.ID",
                                                                               "Current.Age",
                                                                               "Fraction.Genome.Altered",
                                                                               "MSI.Score",
                                                                               "Mutation.Count",
                                                                               "Stage..Highest.Recorded.",
                                                                               "Overall.Survival..Months.",
                                                                               "Overall.Survival.Status",
                                                                               "Tumor.Site..Adrenal.Glands..NLP.",
                                                                               "Tumor.Site..Bone..NLP.",
                                                                               "Tumor.Site..CNS.Brain..NLP.",
                                                                               "Tumor.Site..Intra.Abdominal",
                                                                               "Tumor.Site..Liver..NLP.",
                                                                               "Tumor.Site..Lung..NLP.",
                                                                               "Tumor.Site..Lymph.Node..NLP.",
                                                                               "Tumor.Site..Pleura..NLP.",
                                                                               "Tumor.Site..Reproductive.Organs..NLP.")], 
                                                         by = c("Sample.ID" = "Sample.ID"),
                                                         suffix = c(".cna", ".cbio") ))


dim(clinical_cna_data_for_model)

clinical_cna_data_for_model = clinical_cna_data_for_model[,-grep(".cna",colnames(clinical_cna_data_for_model), ignore.case = T)]
clinical_cna_data_for_model$Cancer.Type[clinical_cna_data_for_model$Cancer.Type == "Colorectal Cancer"] = 1
clinical_cna_data_for_model$Cancer.Type[clinical_cna_data_for_model$Cancer.Type != 1] = 0

#-------------------------------------------------------------------------------

#Training-testing split
training_rows = train_rows(clinical_cna_data_for_model)
training_data = clinical_cna_data_for_model[training_rows,] 
testing_data = clinical_cna_data_for_model[-training_rows,] 

#-------------------------------------------------------------------------------

#Training X
training_x = training_data[, 36:ncol(training_data)] # Select predictors from column 36 onward
training_x$Stage..Highest.Recorded..cbio = factor_2_integer(training_x$Stage..Highest.Recorded..cbio)
training_x$Overall.Survival.Status.cbio = factor_2_integer(training_x$Overall.Survival.Status.cbio)
training_x$Tumor.Site..Adrenal.Glands..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Adrenal.Glands..NLP..cbio)
training_x$Tumor.Site..Bone..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Bone..NLP..cbio)
training_x$Tumor.Site..Bone..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Bone..NLP..cbio)
training_x$Tumor.Site..CNS.Brain..NLP..cbio = factor_2_integer(training_x$Tumor.Site..CNS.Brain..NLP..cbio)
training_x$Tumor.Site..Intra.Abdominal.cbio = factor_2_integer(training_x$Tumor.Site..Intra.Abdominal.cbio)
training_x$Tumor.Site..Liver..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Liver..NLP..cbio)
training_x$Tumor.Site..Lung..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Lung..NLP..cbio)
training_x$Tumor.Site..Lymph.Node..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Lymph.Node..NLP..cbio)
training_x$Tumor.Site..Pleura..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Pleura..NLP..cbio)
training_x$Tumor.Site..Reproductive.Organs..NLP..cbio = factor_2_integer(training_x$Tumor.Site..Reproductive.Organs..NLP..cbio)

#Training Y
training_y = as.factor(training_data$Cancer.Type)   # Convert response variable to factor

#Testing X
testing_x = testing_data[, 36:ncol(testing_data)] # Select predictors from column 36 onward
testing_x$Stage..Highest.Recorded..cbio = factor_2_integer(testing_x$Stage..Highest.Recorded..cbio)
testing_x$Overall.Survival.Status.cbio = factor_2_integer(testing_x$Overall.Survival.Status.cbio)
testing_x$Tumor.Site..Adrenal.Glands..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Adrenal.Glands..NLP..cbio)
testing_x$Tumor.Site..Bone..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Bone..NLP..cbio)
testing_x$Tumor.Site..Bone..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Bone..NLP..cbio)
testing_x$Tumor.Site..CNS.Brain..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..CNS.Brain..NLP..cbio)
testing_x$Tumor.Site..Intra.Abdominal.cbio = factor_2_integer(testing_x$Tumor.Site..Intra.Abdominal.cbio)
testing_x$Tumor.Site..Liver..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Liver..NLP..cbio)
testing_x$Tumor.Site..Lung..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Lung..NLP..cbio)
testing_x$Tumor.Site..Lymph.Node..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Lymph.Node..NLP..cbio)
testing_x$Tumor.Site..Pleura..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Pleura..NLP..cbio)
testing_x$Tumor.Site..Reproductive.Organs..NLP..cbio = factor_2_integer(testing_x$Tumor.Site..Reproductive.Organs..NLP..cbio)

#Testing Y
testing_y = as.factor(testing_data$Cancer.Type)  # True labels

data_for_x_y_combined_models = cbind(Cancer_Type = training_y, training_x)



```


#### Logistic Regression

```{r}

logistic_roc = logistic_regression(training_data = data_for_x_y_combined_models,
                                   testing_x = testing_x,
                                   testing_y = testing_y)  

```

#### Logistic Regression with L1 and L2 penalty (Ridge, Lasso, Elastic Net)

```{r}

ridge_roc = elastic_net_logistic(training_x = training_x,
                                 training_y = training_y,
                                 testing_x = testing_x,
                                 testing_y = testing_y, 
                                 a = 0,
                                 folds = 5,
                                 parallel = TRUE)

lasso_roc = elastic_net_logistic(training_x = training_x,
                                 training_y = training_y,
                                 testing_x = testing_x,
                                 testing_y = testing_y, 
                                 a = 1,
                                 folds = 5,
                                 parallel = TRUE)

elastic_net_roc = elastic_net_logistic(training_x = training_x,
                                 training_y = training_y,
                                 testing_x = testing_x,
                                 testing_y = testing_y, 
                                 a = 1,
                                 folds = 5,
                                 parallel = TRUE)

```

#### Linear discriminant analysis

```{r}

LDA_roc = LDA_function(training_data = data_for_x_y_combined_models,
                       testing_x = testing_x,
                       testing_y = testing_y)

#QDA did not work - "Rank deficiency in 0" error. 

```

#### Random Forest

```{r}

randomforest_roc = random_forest_ranger(training_data = data_for_x_y_combined_models,
                                        training_y = "Cancer_Type",
                                        testing_x = testing_x,
                                        testing_y = testing_y,
                                        num_trees = 2000,
                                        cores = (detectCores()/2))

```

```{r}

xgboost_roc = xgboost_parallel(training_x = training_x,
                               training_y = training_y,
                               testing_x = testing_x,
                               testing_y = testing_y,
                               learning_rate = 0.1,
                               cores = (detectCores()/2),
                               n_tree = 250,
                               tree_depth = 5) 

```

#### ROC Compilation

```{r}

all_models_ROC = data.frame(TPR = ridge_roc$ROC_Data$TPR,
                            FPR = ridge_roc$ROC_Data$FPR,
                            Model = rep("Ridge Logistic", nrow(ridge_roc$ROC_Data)),
                            AUC = rep(ridge_roc$AUC[1], nrow(ridge_roc$ROC_Data)))

all_models_ROC = rbind(all_models_ROC, cbind(lasso_roc$ROC_Data, 
                                             Model = rep("Lasso Logistic", nrow(lasso_roc$ROC_Data)),
                                             AUC = rep(lasso_roc$AUC[1], nrow(lasso_roc$ROC_Data))))

all_models_ROC = rbind(all_models_ROC, cbind(elastic_net_roc$ROC_Data, 
                                             Model = rep("Elastic Net Logistic", nrow(elastic_net_roc$ROC_Data)),
                                             AUC = rep(elastic_net_roc$AUC[1], nrow(elastic_net_roc$ROC_Data))))


all_models_ROC = rbind(all_models_ROC, cbind(logistic_roc$ROC_Data, 
                                             Model = rep("Logistic Regression", nrow(logistic_roc$ROC_Data)),
                                             AUC = rep(logistic_roc$AUC[1], nrow(logistic_roc$ROC_Data))))


all_models_ROC = rbind(all_models_ROC, cbind(LDA_roc$ROC_Data, 
                                             Model = rep("LDA", nrow(LDA_roc$ROC_Data)),
                                             AUC = rep(LDA_roc$AUC[1], nrow(LDA_roc$ROC_Data))))


all_models_ROC = rbind(all_models_ROC, cbind(randomforest_roc$ROC_Data, 
                                             Model = rep("Random Forest", nrow(randomforest_roc$ROC_Data)),
                                             AUC = rep(randomforest_roc$AUC[1], nrow(randomforest_roc$ROC_Data))))

all_models_ROC = rbind(all_models_ROC, cbind(xgboost_roc$ROC_Data, 
                                             Model = rep("XGBoost", nrow(xgboost_roc$ROC_Data)),
                                             AUC = rep(xgboost_roc$AUC[1], nrow(xgboost_roc$ROC_Data))))


all_models_ROC = all_models_ROC %>%
  mutate(Model_AUC = paste(Model, " (AUC = ", round(AUC, 3), ")", sep = ""))

# Plot the ROC curve with AUC in the legend
ggplot(all_models_ROC, aes(x = FPR, y = TPR, color = Model_AUC)) +
  geom_line(size = 1) +
  geom_abline(
    slope = 1, intercept = 0, linetype = "dashed", color = "gray", size = 0.8
  ) +  # Dashed diagonal line for AUC = 0.5
  labs(
    title = "Validation Set ROC Curve",
    x = "False Positive Rate (FPR)",
    y = "True Positive Rate (TPR)",
    color = "Model (AUC)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")


```

## Feature Selection

```{r}

genes_common_to_all_models = Reduce(intersect,
                                    list(ridge_roc$Genes$Genes,
                                         lasso_roc$Genes$Genes,
                                         elastic_net_roc$Genes$Genes,
                                         randomforest_roc$VarImportance$Variable[randomforest_roc$VarImportance$Importance > 10],
                                         xgboost_roc$VarImportance$Feature[xgboost_roc$VarImportance$Gain > 1E-02]))


combined_xgb_rf = inner_join(randomforest_roc$VarImportance[randomforest_roc$VarImportance$Variable %in%
                                                              genes_common_to_all_models, ],
                             xgboost_roc$VarImportance, by = c("Variable" = "Feature"))


df_scaled = combined_xgb_rf %>%
  mutate(Importance_scaled = (Importance - min(Importance)) / (max(Importance) - min(Importance)),
         Gain_scaled = (Gain - min(Gain)) / (max(Gain) - min(Gain)))

# Reshape the scaled data to long format
df_long = df_scaled %>%
  gather(key = "Type", value = "Value", Importance_scaled, Gain_scaled) %>%
  mutate(Model = ifelse(Type == "Importance_scaled", "Random Forest", "XGBoost"))

# Plot
ggplot(df_long, aes(x = reorder(Variable, Value), y = Value, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  scale_fill_manual(values = c("Random Forest" = "springgreen2", "XGBoost" = "violetred2")) +
  labs(
    title = "Scaled Variable Importance | Random Forest vs. XGBoost",
    x = "Variables",
    y = "Scaled Importance/Gain",
    fill = "Model"
  ) +
  theme_minimal()

```



```{r}

cna_imp_features = clinical_cna_data_with_more_markers[,c("Sample.ID",genes_common_to_all_models,"Cancer.Type")]

cna_imp_features_cr_cancer = cna_imp_features[cna_imp_features$Cancer.Type == "Colorectal Cancer",]

imp_genes_cna = cna_imp_features[,2:5]
imp_genes_cna_cr = cna_imp_features_cr_cancer[,2:5]

```

```{r}

cna_imp_features_pivot = as.data.frame(t(imp_genes_cna))

cna_imp_features_pivot$patients_altered[1] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[1,])>0])
cna_imp_features_pivot$patients_altered[2] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[2,])>0])
cna_imp_features_pivot$patients_altered[3] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[3,])>0])
cna_imp_features_pivot$patients_altered[4] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[4,])>0])

cna_imp_features_pivot$patients_altered

ggplot(cna_imp_features_pivot, aes(x = row.names(cna_imp_features_pivot), y = patients_altered, fill = row.names(cna_imp_features_pivot))) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    title = "CNA for the Important Colorectal Cancer Genes in All Patients",
    x = "Genes",
    y = "Number of Patients with Copy Number Alterations"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0.5)) + # Rotate x-axis labels for readability
  scale_fill_discrete(name = "Features") # Add legend for clarity


```

```{r}

cna_imp_features_pivot = as.data.frame(t(imp_genes_cna_cr))

cna_imp_features_pivot$patients_altered[1] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[1,])>0])
cna_imp_features_pivot$patients_altered[2] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[2,])>0])
cna_imp_features_pivot$patients_altered[3] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[3,])>0])
cna_imp_features_pivot$patients_altered[4] = ncol(cna_imp_features_pivot[,abs(cna_imp_features_pivot[4,])>0])

cna_imp_features_pivot$patients_altered

ggplot(cna_imp_features_pivot, aes(x = row.names(cna_imp_features_pivot), y = patients_altered, fill = row.names(cna_imp_features_pivot))) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(
    title = "CNA for the Important Colorectal Cancer Genes in Colorectal Cancer Patients",
    x = "Genes",
    y = "Number of Patients with Copy Number Alterations"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(hjust = 0.5)) + # Rotate x-axis labels for readability
  scale_fill_discrete(name = "Features") # Add legend for clarity


```


```{r}

clinical_cancer_data = clinical_data_cbio[clinical_data_cbio$Cancer.Type == "Colorectal Cancer",]

clinical_cancer_data[,grep("Tumor.Site..", colnames(clinical_cancer_data))]


```



Scaling does NOT improve performance

```{r}

#Scaled training and testing sets

#Training X
training_x = training_data[, 36:ncol(training_data)] # Select predictors from column 36 onward
training_x$Stage..Highest.Recorded..y = as.integer(as.factor(training_x$Stage..Highest.Recorded..y))
training_x$Current.Age = scale(training_x$Current.Age)
training_x$Fraction.Genome.Altered = scale(training_x$Fraction.Genome.Altered)
training_x$MSI.Score = scale(training_x$MSI.Score)
training_x$Mutation.Count = scale(training_x$Mutation.Count)

#Training Y
training_y = as.factor(training_data$Cancer.Type)   # Convert response variable to factor

#Testing X
testing_x = testing_data[, 36:ncol(testing_data)] # Select predictors from column 36 onward
testing_x$Stage..Highest.Recorded..y = as.integer(as.factor(testing_x$Stage..Highest.Recorded..y))
testing_x$Current.Age = scale(testing_x$Current.Age)
testing_x$Fraction.Genome.Altered = scale(testing_x$Fraction.Genome.Altered)
testing_x$MSI.Score = scale(testing_x$MSI.Score)
testing_x$Mutation.Count = scale(testing_x$Mutation.Count)

#Testing Y
testing_y = as.factor(testing_data$Cancer.Type)  # True labels

data_for_x_y_combined_models = cbind(Cancer_Type = training_y, training_x)

unique(clinical_cna_data$Sample.Class)
#clinical_cna_data

```





```{r}

```


Scale the data above

Look at the data below


```{r}

data_timeline_tumor_markers

data_timeline_diagnosis
data_sv

na_cleanup_columns(data_mutations)

plot = ggplot(data_timeline_tumor_markers, aes(x = START_DATE, y = RESULT, color = TEST, group = PATIENT_ID)) +
  geom_line() +
  labs(x = "Start Date", y = "Result (ng/mL)", title = "Test Results Over Time") +
  theme_minimal()


ggsave("/Users/ali/Desktop/NYU/3rd Semester/plot.png", plot, bg = "white")



```





























