# Final Project (BIOL-GA 2031): Colorectal Cancer

**Muhammad Ali** \| N18480700

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Loading the libraries

```{r}

suppressPackageStartupMessages({
library(tidyr)
library(dplyr)
library(pROC)
library(randomForest)
library(ggplot2)
library(MASS)
library(e1071)
library(glmnet)
})


```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Loading the data

```{r}

#Demographic data
data_clinical_patients = read.delim("/Users/ali/Downloads/msk_chord_2024/data_clinical_patient.txt", header = T)

#Has cancer type information
data_clinical_sample = read.delim("/Users/ali/Downloads/msk_chord_2024/data_clinical_sample.txt", header = T)

#Profile_description: Putative copy-number from GISTIC 2.0. Values: -2 = homozygous deletion; -1 = hemizygous deletion; 0 = neutral / no change; 1 = gain; 2 = high level amplification.
#Profile_name: Putative copy-number alterations from GISTIC
data_cna = read.delim("/Users/ali/Downloads/msk_chord_2024/data_cna.txt", header = T)
data_cna = data.frame(t(data_cna))
colnames(data_cna) = data_cna[1,]
data_cna = data_cna[-1,]
row.names(data_cna) = gsub("\\.","-",row.names(data_cna))
data_cna[] <- lapply(data_cna, as.integer)
data_cna$Sample_ID = row.names(data_cna)

#Some sort of mutation data
data_gene_panel_matrix = read.delim("/Users/ali/Downloads/msk_chord_2024/data_gene_panel_matrix.txt", header = T)

#Some sort of mutation data
data_mutations = read.delim("/Users/ali/Downloads/msk_chord_2024/data_mutations.txt", header = T)

#Profile_name: Structural Variants
#Profile_description: Structural Variant Data.
data_sv = read.delim("/Users/ali/Downloads/msk_chord_2024/data_sv.txt")

#Some temporal data about cancer
data_timeline_cancer_presence = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_cancer_presence.txt")

#Some temporal diagnosis data
data_timeline_diagnosis = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_diagnosis.txt")

#Has something called the Gleason score
data_timeline_gleason = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_gleason.txt")

#Something about MMR presence or absence. 
data_timeline_mmr = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_mmr.txt")

#Something along the lines of a PDL1 positivity test
data_timeline_pdl1 = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_pdl1.txt")

#Something about ECOG
data_timeline_performance_status = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_performance_status.txt")

#Data about prior medication
data_timeline_prior_meds = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_prior_meds.txt")

#Data about progression
data_timeline_progression = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_progression.txt")

#Data about raditaion treatment
data_timeline_radiation = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_radiation.txt")

#Something about sample acquisition and sample IDs
data_timeline_specimen_surgery = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_specimen_surgery.txt")

#Something about sample ID
data_timeline_specimen = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_specimen.txt")

#Something about whether the sample was collected or the procedure. 
data_timeline_surgery = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_surgery.txt")

#Data about treatment type and start and stop date
data_timeline_treatment = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_treatment.txt")

#Probably data about tests that measure different biomarker levels
data_timeline_tumor_markers = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_tumor_markers.txt")

#Data about tumor location
data_timeline_tumor_sites = read.delim("/Users/ali/Downloads/msk_chord_2024/data_timeline_tumor_sites.txt")

#-------------------------------------------------------------------------------

#Case Data

#All samples
cases_all = data.frame(Sample = t(read.delim("/Users/ali/Downloads/msk_chord_2024/case_lists/cases_all.txt", header = F)))

#Samples with CNA data
cases_cna = data.frame(Sample_CNA = t(read.delim("/Users/ali/Downloads/msk_chord_2024/case_lists/cases_cna.txt", 
                                                 header = F)))

#Samples with mutation and CNA data
cases_cnaseq = data.frame(Sample_CNA_mutation = t(read.delim("/Users/ali/Downloads/msk_chord_2024/case_lists/cases_cnaseq.txt",
                                                             header = F)))

#Samples with mutation data
cases_sequenced = data.frame(Sample_mutation = t(read.delim("/Users/ali/Downloads/msk_chord_2024/case_lists/cases_sequenced.txt",
                                                            header = F)))

#All samples with SV data
cases_sv = data.frame(Sample_SV = t(read.delim("/Users/ali/Downloads/msk_chord_2024/case_lists/cases_sv.txt",
                                               header = F)))


#-------------------------------------------------------------------------------

#Clinical data directly from cBioPortal

clinical_data_cbio = read.delim("/Users/ali/Downloads/msk_chord_2024_clinical_data-2.tsv")


```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

### Custom functions

```{r}

#Removes columns that have NAs in them - machine learning models cannot handle NAs; difficult to impute values when placeholders such as "0" might be used to singify a disease state. Imputation is difficult without making big assumptions, especially since this is beca use some test may not have been performed for certain patients. Therefore, I removed those columns entirely. 
na_cleanup_columns = function(dataframe)
{
  return(dataframe[,colSums(is.na(dataframe)) == 0])
}

#-------------------------------------------------------------------------------

#Creates training rows
train_rows = function(x)
{
  sample(1:nrow(x),nrow(x)*(2/3), replace = F)
}


```

\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_\_

Analyze genomic and clinical data differently and then integrate at the end

```{r}

first_join = inner_join(data_clinical_patients, data_clinical_sample, by = c("PATIENT_ID" = "PATIENT_ID"))

second_join = inner_join(first_join, data_gene_panel_matrix, by = c("SAMPLE_ID" = "SAMPLE_ID"))

third_join = inner_join(second_join, data_timeline_diagnosis, by = c("PATIENT_ID" = "PATIENT_ID"))

fourth_join = inner_join(third_join, data_timeline_specimen_surgery, by = c("PATIENT_ID" = "PATIENT_ID"))

fifth_join = inner_join(fourth_join, data_timeline_specimen, by = c("PATIENT_ID" = "PATIENT_ID"))

fifth_join







#Increases the number of rows
data_mutations
data_timeline_cancer_presence
data_timeline_performance_status #(Only need start date and ECOG - higher ECOG scores mean poorer quality of life)
data_timeline_progression #Can take data from procedure type and of progression
data_timeline_radiation #Treatment start data of radiation therapy
data_timeline_surgery #The start date for whether surgery was performed or sample was obtained 
data_timeline_treatment 
data_timeline_tumor_markers
data_timeline_tumor_sites

#reduces the number of rows
data_timeline_gleason
data_sv
data_timeline_mmr
data_timeline_pdl1
data_timeline_prior_meds

```

**Training & testing split**

```{r}

set.seed(1010)

clinical_cna_data = na_cleanup_columns(inner_join(clinical_data_cbio, data_cna, by = c("Sample.ID" = "Sample_ID")))

clinical_cna_data_for_model = clinical_cna_data

clinical_cna_data_for_model$Cancer.Type[clinical_cna_data_for_model$Cancer.Type == "Colorectal Cancer"] = 1
clinical_cna_data_for_model$Cancer.Type[clinical_cna_data_for_model$Cancer.Type != 1] = 0

#-------------------------------------------------------------------------------

#Training-testing split
training_rows = train_rows(clinical_cna_data_for_model)
training_data = clinical_cna_data_for_model[training_rows,] 
testing_data = clinical_cna_data_for_model[-training_rows,] 

#-------------------------------------------------------------------------------

x = training_data[, 36:ncol(training_data)] # Select predictors from column 36 onward
y = as.factor(training_data$Cancer.Type)   # Convert response variable to factor
true_labels = testing_data$Cancer.Type  # True labels

data_for_x_y_combined_models = cbind(Cancer_Type = y,x)

```


#### Lasso logistic regression

```{r}

# Fit the L1-penalized logistic regression (Lasso)
cv_lasso_logistic = cv.glmnet(as.matrix(x), y, family = "binomial", alpha = 1, nfolds = 5)

testing_x_regularized_logisitic = as.matrix(testing_data[, 36:ncol(testing_data)])
lasso_probs = predict(cv_lasso_logistic, 
                      newx = testing_x_regularized_logisitic,
                      type = "response",
                      s=cv_lasso_logistic$lambda.1se)

# Calculate AUC
roc_lasso = pROC::roc(testing_data$Cancer.Type, lasso_probs)

# Plot ROC Curve
roc_data = data.frame(TPR = roc_lasso$sensitivities, 
                      FPR = 1 - roc_lasso$specificities)

ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | Lasso Logistic Regression | AUC =", round(roc_lasso$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()

```

#### Ridge logistic regression

```{r}

# Fit the L1-penalized logistic regression (Lasso)
cv_lasso_logistic = cv.glmnet(as.matrix(x), y, family = "binomial", alpha = 0, nfolds = 5)

testing_x_regularized_logisitic = as.matrix(testing_data[, 36:ncol(testing_data)])
lasso_probs = predict(cv_lasso_logistic, 
                      newx = testing_x_regularized_logisitic,
                      type = "response",
                      s=cv_lasso_logistic$lambda.1se)

# Calculate AUC
roc_lasso = pROC::roc(testing_data$Cancer.Type, lasso_probs)

# Plot ROC Curve
roc_data = data.frame(TPR = roc_lasso$sensitivities, 
                      FPR = 1 - roc_lasso$specificities)

ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | Ridge Logistic Regression | AUC =", round(roc_lasso$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()

```

#### Elastic-net logistic regression

```{r}

# Fit the L1-penalized logistic regression (Lasso)
cv_lasso_logistic = cv.glmnet(as.matrix(x), y, family = "binomial", alpha = 0.5, nfolds = 5)

testing_x_regularized_logisitic = as.matrix(testing_data[, 36:ncol(testing_data)])
lasso_probs = predict(cv_lasso_logistic, 
                      newx = testing_x_regularized_logisitic,
                      type = "response",
                      s=cv_lasso_logistic$lambda.1se)

# Calculate AUC
roc_lasso = pROC::roc(testing_data$Cancer.Type, lasso_probs)

# Plot ROC Curve
roc_data = data.frame(TPR = roc_lasso$sensitivities, 
                      FPR = 1 - roc_lasso$specificities)

ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "blue", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | Elastic-net Logistic Regression | AUC =", round(roc_lasso$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()

```

```{r}

coefficients_glmnet = as.data.frame(as.matrix(coef(cv_lasso_logistic, s = cv_lasso_logistic$lambda.1se)))

genes_selected = rownames(coefficients_glmnet)[coefficients_glmnet$s1 != 0][-1] #removing the intercept

```


#### Logistic Regression

```{r}

# Fit the logistic regression model
logistic_model = glm(Cancer_Type ~ ., 
                     data = data_for_x_y_combined_models, 
                     family = "binomial")

#summary(logistic_model)

predictions = predict(logistic_model, 
                      newdata = testing_data[, 36:ncol(testing_data)],
                      type = "response")

# Convert probabilities to class labels if needed (e.g., using a threshold of 0.5)
predicted_classes = ifelse(predictions > 0.85, 1, 0)

mean(predicted_classes==testing_data$Cancer.Type)

roc_curve = pROC::roc(true_labels, predictions)

# Extract ROC curve data
roc_data = data.frame(TPR = roc_curve$sensitivities,  # True positive rate
                      FPR = 1 - roc_curve$specificities)  # False positive rate

# Create the ROC curve plot
ggplot(roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "skyblue", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +  # Diagonal line
  labs(title = paste("ROC Curve |","AUC =", round(roc_curve$auc, 3)), 
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()

```

#### Linear discriminant analysis

```{r}

# Fit LDA model
lda_model = lda(Cancer_Type ~ ., data = data_for_x_y_combined_models)

# Make predictions
lda_predictions = predict(lda_model, newdata = testing_data[, 36:ncol(testing_data)])

# Predicted probabilities for ROC curve
lda_probs = lda_predictions$posterior[, 2]  # Probabilities for class "1"

# Calculate AUC
lda_roc = pROC::roc(testing_data$Cancer.Type, lda_probs)

# Plot ROC curve
lda_roc_data = data.frame(TPR = lda_roc$sensitivities, 
                          FPR = 1 - lda_roc$specificities)

ggplot(lda_roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "skyblue", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | LDA | AUC =", round(lda_roc$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()


```


```{r}

#degree parameter used for polynomial kernels to set the degree of the polynomial.

# Fit SVM model
svm_model <- svm(Cancer_Type ~ ., 
                 data = data_for_x_y_combined_models, 
                 probability = TRUE,
                 kernel = "radial",
                 cost = 1)


# Make predictions
svm_probs <- attr(predict(svm_model, newdata = testing_data[, 36:ncol(testing_data)], 
                          probability = TRUE), "probabilities")[, 2]

# Calculate AUC
svm_roc <- pROC::roc(testing_data$Cancer.Type, svm_probs)
print(svm_roc$auc)

# Plot ROC curve
svm_roc_data <- data.frame(TPR = svm_roc$sensitivities, 
                           FPR = 1 - svm_roc$specificities)
ggplot(svm_roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "purple", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | SVM | AUC =", round(svm_roc$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()


```



#-------------------------------------------------------------------------------


RANDOM FORESTS taking forever

```{r}

mtry_value = floor(sqrt(ncol(data_for_x_y_combined_models)))

rf_model = randomForest(x = x,
                        y = y,
                        mtry = mtry_value, 
                        ntree = 500,
                        importance = TRUE)

# Make predictions
rf_probs = predict(rf_model, newdata = testing_data[, 36:ncol(testing_data)], type = "prob")[, 2]

# Calculate AUC
rf_roc = pROC::roc(testing_data$Cancer.Type, rf_probs)

# Plot ROC curve
rf_roc_data = data.frame(TPR = rf_roc$sensitivities, 
                         FPR = 1 - rf_roc$specificities)

ggplot(rf_roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "darkorange", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | Random Forest | AUC =", round(rf_roc$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()

```


QDA NOT WORKING for CNA data

```{r}

# Fit QDA model
qda_model = qda(Cancer_Type ~ ., data = data_for_x_y_combined_models)

# Make predictions
qda_predictions = predict(qda_model, 
                           newdata = testing_data[, 36:ncol(testing_data)])

# Predicted probabilities for ROC curve
qda_probs <- qda_predictions$posterior[, 2]  # Probabilities for class "1"

# Calculate AUC
qda_roc <- pROC::roc(testing_data$Cancer.Type, qda_probs)
print(qda_roc$auc)

# Plot ROC curve
qda_roc_data <- data.frame(TPR = qda_roc$sensitivities, 
                           FPR = 1 - qda_roc$specificities)
ggplot(qda_roc_data, aes(x = FPR, y = TPR)) +
  geom_line(color = "forestgreen", size = 1) +
  geom_abline(linetype = "dashed", color = "red") +
  labs(title = paste("ROC Curve | QDA | AUC =", round(qda_roc$auc, 3)),
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_minimal()


```





### Predictive pipeline


#### Discriminant Analysis

#### Random Forests

#### SVM

#### Neural Networks
